Landau quantization of nearly degenerate bands, \\ and full symmetry classification of avoided Landau-level crossings

### 2019-01-31 paper.tex
submitted to PRX

\begin{abstract}
Semiclassical quantization rules compactly describe the energy dispersion of Landau levels, and are predictive of quantum oscillations in transport and thermodynamic quantities. Such rules –- as formulated by Onsager, Lifshitz and Roth –-
apply when the spin-orbit interaction dominates over the Zeeman interaction (or vice versa), but does not generally apply  when the two interactions are comparable in strength. In this work, we present a generalized quantization rule which treats the spin-orbit and Zeeman interactions on equal footing, and therefore has wider applicability to spin-orbit-coupled materials lacking a spatial inversion center, or having magnetic order. More generally, our rule describes the Landau quantization of any number of nearly degenerate energy bands -- in any symmetry class. The resultant Landau-level spectrum is generically non-equidistant but may contain \blue{spin (or pseudospin)} degeneracies. To tune to such degeneracies in the absence of crystalline point-group symmetries, three real parameters are needed. We have exhaustively identified all symmetry classes of cyclotron orbits for which this number is reduced from three, thus establishing symmetry-enforced`non-crossing rules' for Landau levels.
In particular, only one parameter is needed {in the presence of} spatial rotation {or inversion}; this single parameter may be the magnitude or orientation of the field. Signatures of single-parameter tunability include (i) a smooth crossover between period-doubled and -undoubled quantum oscillations in the low-temperature Shubnikov-de Haas effect, as well as (ii) `magic-angle' magnetoresistance oscillations. We demonstrate the utility of our quantization rule, as well as the tunability of Landau-level degeneracies, for the Rashba-Dresselhaus two-dimensional electron gas -- subject to an arbitrarily oriented magnetic field.
\end{abstract}
\date{\today}

@@

\section{Introduction}
In the semiclassical theory of a Bloch electron in a magnetic field,  quantization rules not  only serve as a compact description of the  energy dispersion of Landau levels, they also form the basis for a quantitative theory of quantum oscillations in transport\cite{SdH} and thermodynamic\cite{dHvA} quantities.  The original formulation of the rule  -- by Onsager\cite{Onsager} and Lifshitz\cite{lifshitz_kosevich,lifshitz_kosevich_jetp} -- neglects the effect of spin-orbit coupling. While subsequent generalizations by Roth,
\cite{rotheffham,rothmag} Mikitik\cite{Mikitik_quantizationrule},
and the present authors\cite{topoferm,100p} account for the spin-orbit effect, such quantization rules apply only when the spin-orbit splitting of energy bands is much stronger (or much weaker) than the Zeeman interaction.

A quantization rule that treats the spin-orbit and Zeeman interactions on equal footing has yet to be formulated. Such a rule would be especially relevant to spin-orbit-coupled materials lacking a spatial inversion center, or having magnetic order. The energy bands of these materials carry a spin-half degree of freedom, and the spin degeneracy of energy bands is weakly lifted by spin-orbit coupling or magnetic ordering. A prototypical example is given by  a two-dimensional electron gas (2DEG) with Rashba spin-orbit coupling:
\begin{equation}
H_R=\frac{{\hbar^2} k^2}{2m}+\hbar\alpha  (k_{x}\sigma_{y}-k_{y}\sigma_{x}).\label{eq:Rashba-Hamiltonian}
\end{equation}
Given a magnetic field in the $-z$ direction, and focusing on energies much greater than $m\alpha^2$, the cyclotron orbits [illustrated in \fig{fig:orbits}(a)]   are split in $\bk$-space with a differential area  $\delta S$ that is much less than their average area $S$:\footnote{The expressions for $\delta S$ and $S$ in \q{deltaSvsS} are asymptotically valid for $E{\gg}m\alpha^2$.}
\e{\delta S = 4\pi m \alpha k_E/\hbar\ll S =\pi k_E^2; \;\; k_{E}:=\f{\sqrt{2m E}}{\hbar}.\label{deltaSvsS}}
$S$ is equivalently the area of the constant-energy contour in the absence of spin-orbit coupling ($\alpha{=}0$).

@@

Our work presents a generalized rule to describe the Landau quantization of nearly-degenerate energy bands, for which  spin-orbit-split bands is but one example. \blue{By the \textit{near-degeneracy} condition, we generally mean that
the ratio of differential to average area is small: $|\delta S/S|{\ll}1$. Our proposed rule relies also on a second condition -- that $S$ is much larger than the inverse square of the magnetic length $l{=}\sqrt{\hbar/eB}$; this is the standard semiclassical condition in the Onsager-Lifshitz-Roth theory.} By treating \textit{both} $|\delta S/S|$ and $1/l^2|S|$   as small parameters with finite ratio $l^2|\delta S|$, our perspective diverges from and generalizes \textit{single-parameter} semiclassical theories.\cite{kohn_effham,blount_effham,rotheffham,wannier_fredkin,fischbeck_review,Mikitik_quantizationrule,topoferm,100p,gao_zero-field_2017}

The inverse ratio $1/l^2|\delta S|$ measures the hybridization of nearly-degenerate orbits by the Zeeman interaction. In the regime $l^2|\delta S|{\gg}1$, the Onsager-Lifshitz{-Roth} quantization rules apply independently to  two concentric (but decoupled) orbits. In the opposite regime $l^2|\delta S|{\ll}1$, the spin-orbit splitting of energies is negligible, and we may then apply the known quantization rule for exactly-degenerate bands\cite{rotheffham,rothmag,topoferm,100p,Mikitik_quantizationrule}. Our unifying rule provides the smooth interpolation between these two known regimes. {Equivalently stated, we have extended the applicability of quantization rules to the intermediate regime ($l^2|\delta S|{\sim}1$) where the Zeeman interaction is comparable with the spin-orbit interaction; previously, such rules were known only when one interaction dominates over the other. This extension allows  to formulate semiclassical quantization rules for spin-orbit coupled materials in \textit{any} symmetry class -- with or without spatial inversion symmetry, with or without time-reversal symmetry. The unifying rule is presented in \s{sec:qtznrules} for nearly-degenerate bands with a spin\blue{-half (or pseudospin-half)} degree of freedom. As elaborated in \s{sec:discussion}, the most general formulation of our rule is applicable  to any number of nearly-degenerate bands of any symmetry.}

In \s{sec:Rashba}, we demonstrate the utility of this rule for the Rashba{-Dresselhaus} 2DEG with an arbitrarily-oriented magnetic field. A sufficiently strong in-plane field not only moves the Dirac point but tilts the Dirac cone over -- into a type-II\cite{soluyanov_type-ii_2015, muechler_tilted_2016, bergholtz_topology_2015} Dirac point. In \s{sec:inplanezeeman}, we investigate the Landau-Zener tunneling dynamics that occur in the vicinity of a II-Dirac point; in particular, we will {demonstrate that our quantization rule can account for quantum tunneling between nearly-degenerate bands -- a phenomenon known as interband breakdown\cite{kaganov_coherent_1983,slutskin_dynamics_1968,AALG,100p}. We will also} critically evaluate a recent claim\cite{obrien_magnetic_2016} of perfect Klein tunneling at the energy of the Dirac point -- \blue{we find that such tunneling is never perfect with a proper account of the Zeeman interaction.}

@@

In this section, we apply our rule to the Landau levels of the Rashba model [cf.\ \q{eq:Rashba-Hamiltonian}], for which analytic, closed-form solutions exist for verification.\cite{bychkov_oscillatory_1984}

For simplicity, we assume that the Pauli matrices in \q{eq:Rashba-Hamiltonian} correspond to spin: $s_i{=}\hbar \sigma_i/2$. Let us evaluate the effective Hamiltonian $\calh(\bk)$ along the cyclotron trajectory $\frako_0$ with radius $k$. In the energy basis of the Rashba Hamiltonian $H_R$ [cf.\ \q{eq:Rashba-Hamiltonian}], 
\e{\calh(\bk)=-\hbar\alpha k\tau_z{+}\f{g_{s\perp}}{2}\mu_{\sma{B}}B\tau_x {-} \f{eB\hbar}{2m} (\tau_x-I) \epsilon_{\alpha\beta} k_{\alpha}\nabla_{\beta}\theta,\label{rashbaeffham}}
where $\tau_i$ are a new set of Pauli matrices with $\tau_z{=}{+} 1$ and ${-}1$ corresponding to the outer and inner orbits, respectively. $e^{i\theta_{\bk}}{=}(ik_x{-}k_y)/k$ is the angle of the in-plane spin-splitting field $(-\hbar\alpha k_y,\hbar\alpha k_x,0)$ of $H_R$ [cf.\ \q{eq:Rashba-Hamiltonian}], which winds once over $\frako_0$; the spin of our basis vectors likewise winds, in parallel (or anti-parallel) alignment with this field (see Fig. (\ref{fig:orbits})(b)). Terms proportional to the rate of spin rotation ($\nabk \theta$) originate from the non-Abelian Berry connection. The first term of  $\calh$ corresponds to the spin-orbit-induced splitting of the energy-degenerate bands;
the remaining terms may be interpreted as $B$-induced deformations of the bands. The orbital moment term $B M_z$, which generally couples the two-band subspace (of interest) to all other bands, is absorbed into the Zeeman spin interaction with a phenomenological g-factor $g_{s\perp}$.\footnote{For simplicity, we neglect here off-diagonal elements of $M_z$ in the basis of $S_z=\pm 1/2$, with $\vec{z}$ the direction of the field.} It is well-known that the Zeeman term tends to polarize spin in the direction of $\bB$; remarkably, the non-Abelian Berry term  also has a polarizing effect, but with opposite sign. These two effects completely cancel for free electrons (with $g_{s\perp}=2$ and $m$ equal to the free-electron mass $m_0$), but not generically in band systems. 
%is absent in any simplified two-band model.

@@

In the weak-hybridization regime: $l^2\delta S \gg \pi|g_{s\perp} m/m_0{-}2|$, we may expand \q{eq:Rashba-lambda} as a Laurent series in $B$:
\e{\lambda_{\pm}= \pm \f{l^2\delta S}{2} +\pi \pm \f{\pi^2}{4}\f{(g_{s\perp}m/m_0-2)^2}{l^2\delta S} + O(B^2). \label{lambdasmallB}}
\q{eq:rule}, combined with the first two terms on the right-hand-side of \q{lambdasmallB}, may alternatively be derived from  the standard Onsager-Lifshitz rules for two independent orbits, with a common $\pi$ Berry-phase correction for each orbit:\footnote{The intra-band Zeeman and orbital-moment couplings vanish due spacetime-inversion symmetry\cite{topoferm}.} $l^2(S{\pm}\delta S)/2\pi {\in}\Z$. Higher-order terms in \q{lambdasmallB} reflect the field-induced hybridization of the spin-orbit-split orbits. Eq.\ (\ref{lambdasmallB}) has been partially obtained by Mineev for the Rashba model;\cite{mineev_haas--van_2005} he missed the second term in $(g_{s\perp}m/m_0-2)$ [cf.\ \q{lambdasmallB}], due to his neglect of the interband Berry connection.

@@

Our quantization rule [\q{eq:rule} with \q{eq:Rashba-lambda}] may be compared with the  exact Landau-level spectrum of the Peierls-substituted Hamiltonian $H_R(\bK)$ (plus the Zeeman coupling to spin).  The exact levels\cite{bychkov_oscillatory_1984} are given implicitly by
\e{
&n_{\pm}=\frac{l^2S}{2\pi}+\frac{l^2\delta S}{16\pi}\f{\delta S}{S}\lin 
&\pm\f1{2}\sqrt{\bigg(\frac{l^2\delta S}{8\pi}\f{\delta S}{S}\bigg)^2+(l^2\delta S)^2+\pi^2\bigg(\f{g_{s\perp}m}{m_0}-2\bigg)^2},\label{eq:Rashba-exact}}
with the functional forms of $\delta S(E)$ and $S(E)$ provided in \q{deltaSvsS};  $n_{+}{\ge} 1$ and $n_-{\ge} 0$ here are integer-valued labels for two distinct sets of levels. \blue{In the near-degenerate ($\delta S/S\ll 1$) and semiclassical ($1/l^2S \ll 1$) regime, with finite $l^2\delta S$,}  the two terms in \q{eq:Rashba-exact} involving $\delta S/S$  may be neglected, and \q{eq:Rashba-exact} reduces to our quantization rule with $\lambda_{\pm}$ given by \q{eq:Rashba-lambda}.

\blue{With the further inclusion of the Dresselhaus spin-orbit interaction in \q{eq:Rashba-Hamiltonian}, no closed-form analytic solution (of the Landau levels) is known. We may anyway verify the validity of our approximation scheme  -- by comparison with the Landau levels obtained from numerical exact diagonalization; this  is carried out in  \app{app:approximatevsexact}.}

%The just-stated inequality is equivalent to an upper bound on the Landau-level index: $n_{\pm}{\ll} (S/\delta S)^2$, which is distinct from the lower bound $n_{\pm}{\gg} 1$ dictated by the usual semiclassical condition.

%In the near-degenerate regime ($E{\gg}m\alpha^2$),

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{RZ.png}
\caption{(a) illustrates the cyclotron orbits of the Rashba 2DEG in a \blue{tilted} magnetic field. The inset illustrates an enlarged view of near II-Dirac point. (b) illustrates the cyclotron orbit at the energy of the II-Dirac point; \blue{a Klein-tunneling electron will follow a trajectory that resembles the edge of a M\"obius strip.} The orbit is drawn over the  spin-splitting field ($d_x,d_y$) of the degeneracy-lifting Hamiltonian $\delta\var$ [cf.\ \q{eq:RZ-Hamiltonian}]. The wavefunction along the inner (resp.\ outer) component of the \blue{M\"obius-strip} orbit has its pseudospin aligned parallel (resp.\ anti-parallel) to the spin-splitting field. (c) schematically illustrates  Landau levels near the II-Dirac point (in the weak-hybridization regime). The dashed line represents the Onsager-Lifshitz-Roth quantization of the \blue{M\"obius-strip} orbit [cf.\ Eq.\ (\ref{trefoilrule})], and the solid line incorporates the first-order correction [cf.\ Eq.\ (\ref{RZfirstorderE})] due to tunneling. (d) \blue{The black solid line is the plot of $\delta E_n$ [cf. Eq. (\ref{RZfirstorderE})], which is the energetic correction due to deviations from perfect Klein tunneling; for comparison, the red dashed line shows the numerically-exact deviations obtained from exact diagonalization.} This comparison is made for the Landau levels closest to 0.45eV and with even Landau-level index $n$; for the parameters we have chosen ($m{=}0.076m_0$, $\alpha{=} 1.5\times10^{5}$cm/s, $Z_\parallel {=}$1meV and $g_{s\perp}=2$),  the II-Dirac point lies at 0.5eV. 
%Out-of-plane Zeeman coupling is negligibly weak.
\label{fig:RZ}}
\end{figure}

@@

As alluded to in \s{sec:qtznrules}, the adiabatic limit of two independent orbits does not exist if the two spin-orbit-split orbits exactly intersect at a II-Dirac point. To investigate this failure of adiabaticity, let us study the Landau levels at energies close to the II-Dirac point ($\epsilon_0$). We focus first on the regime where, on average (over $\frako_0$), the energy splitting $|\delta \var_+{-}\delta \var_-|$ [cf.\ \q{eq:RZ-Hamiltonian}] is much greater than $B_\perp|\calm_{\sma{+-}}|$. Equivalently, we are in the regime where $\hbar\alpha k_{\sma{E}}{\gg}\var_c^\perp$.\footnote{The orbit-averaged energy splitting is of order $\alpha k_{\sma{E}}$ for energies near the Dirac point. $B_\perp|\calm_{\sma{+-}}|$ is simply the Berry contribution to the Zeeman interaction, and has the form of the right-most term in \q{rashbaeffham}; this term is of order $\var_c$.} This implies that the split orbits do not appreciably hybridize -- except in the vicinity of the II-Dirac point where orbits (nearly) touch.  Near this touching point (where $t{=}0$), the dynamics of a spinor electron is described by linearizing $\H$ [cf.\ \q{calHRZ}] with respect to $t$:
\e{
\calh = -\hbar\alpha k_{E}\omega^{\sma{\perp}}_c t\,\sigma_x+(Z_\parallel-\alpha k_{E})\sigma_y-Z_\perp\sigma_z,\label{effhamIIDirac}
}
The eigenbasis of $\sigma_x$ corresponds to two diabatic levels with characteristic slope $v_d{:}{=}\hbar\alpha k_E\omega^{\sma{\perp}}_c$; near $t{=}0$, the diabatic levels anticross  due to the hybridization terms proportional to $\sigma_{y,z}$. Generally, the probability of tunneling across the hybridization-induced barrier is  $\rho^2{=}\exp(-2\pi\barmu)$, with $\barmu{=}{E_g}^2/{2v_d \hbar}$ and $E_g$ the energy gap at the center of the anticrossing\cite{wittig_landauzener_2005,lifshitz_e.m._quantum_1991}; in this context,
\e{\barmu=\frac{Z_\perp^2+(Z_\parallel-\alpha k_{E})^2}{2\hbar\alpha k_{E}\var^{\sma{\perp}}_c}, \label{mu1}}
with $2\hbar\alpha k_E$ the spin-orbit splitting at energy $E$ [cf.\ \fig{fig:orbits}(a)]. In the absence of the out-of-plane Zeeman coupling ($Z_\perp{=}0$), $\bar{\mu}$ would vanish at the energy of the II-Dirac point -- leading to unit-probability Klein tunneling, as was first proposed by Ref. \onlinecite{obrien_magnetic_2016} in a different model of the II-Dirac point. However, with a proper account of $Z_\perp$, we find instead that $\bar{\mu}$ has a nonvanishing minimum: $\bar{\mu}_{\sma{\text{min}}}{:}{=} (g_{s\perp} m/m_0)(Z_\perp/4\hbar\alpha k_{E})$, which is a product of the dimensionless effective mass and the ratio of the out-of-plane Zeeman splitting over the spin-orbit splitting. \blue{This minimum may be significant for semiconductor heterostructures (e.g.,  $g_{s\perp}m/m_0{\sim} 1$ in InAs heterostructures\cite{pakmehr_g-factor_2015}) and  for heavy-fermion systems.}

@@

\section{Proof of nearly-degenerate quantization rule\label{app:quantizationruleproof}}

>>>>

\begin{abstract}
Semiclassical quantization rules compactly describe the energy dispersion of Landau levels, and are predictive of quantum oscillations in transport and thermodynamic quantities. Such rules -- as formulated by Onsager, Lifshitz and Roth --
apply when the spin-orbit interaction dominates over the Zeeman interaction (or vice versa), but does not generally apply when the two interactions are comparable in strength. In this work, we present a generalized quantization rule which treats the spin-orbit and Zeeman interactions on equal footing, and therefore has wider applicability to spin-orbit-coupled materials lacking a spatial inversion center, or having magnetic order. More generally, our rule describes the Landau quantization of any number of nearly degenerate energy bands -- in any symmetry class. The resultant Landau-level spectrum is generically non-equidistant but may contain spin (or pseudospin) degeneracies. To tune to such degeneracies in the absence of crystalline point-group symmetries, three real parameters are needed. We have exhaustively identified all symmetry classes of cyclotron orbits for which this number is reduced from three, thus establishing symmetry-enforced `non-crossing rules' for Landau levels.
In particular, only one parameter is needed {in the presence of} spatial rotation {or inversion}; this single parameter may be the magnitude or orientation of the field. Signatures of single-parameter tunability include (i) a smooth crossover between period-doubled and -undoubled quantum oscillations in the low-temperature Shubnikov-de Haas effect, as well as (ii) `magic-angle' magnetoresistance oscillations. We demonstrate the utility of our quantization rule, as well as the tunability of Landau-level degeneracies, for the Rashba-Dresselhaus two-dimensional electron gas -- subject to an arbitrarily oriented magnetic field.
\end{abstract}
\date{\today}

@@

\section{Introduction}
In the semiclassical theory of a Bloch electron in a magnetic field,  quantization rules not  only serve as a compact description of the  energy dispersion of Landau levels, they also form the basis for a quantitative theory of quantum oscillations in transport\cite{SdH} and thermodynamic\cite{dHvA} quantities.  The original formulation of the rule -- by Onsager\cite{Onsager} and Lifshitz\cite{lifshitz_kosevich,lifshitz_kosevich_jetp} -- neglects the effect of spin-orbit coupling. While subsequent generalizations by Roth\cite{rotheffham,rothmag}, Mikitik\cite{Mikitik_quantizationrule}, and the present authors\cite{topoferm,100p} account for the spin-orbit effect, such quantization rules apply only when the spin-orbit splitting of energy bands is much stronger (or much weaker) than the Zeeman interaction.

A quantization rule that treats the spin-orbit and Zeeman interactions on equal footing has yet to be formulated. Such a rule would be especially relevant to spin-orbit-coupled materials lacking a spatial inversion center, or having magnetic order. The energy bands of these materials carry a spin-half degree of freedom, and the spin degeneracy of energy bands is weakly lifted by spin-orbit coupling or magnetic ordering. A prototypical example is given by  a two-dimensional electron gas (2DEG) with Rashba spin-orbit coupling:
\begin{equation}
H_R=\frac{{\hbar^2} k^2}{2m}+\hbar\alpha  (k_{x}\sigma_{y}-k_{y}\sigma_{x}).\label{eq:Rashba-Hamiltonian}
\end{equation}
Given a magnetic field in the $-z$ direction, and focusing on energies much greater than $m\alpha^2$, the cyclotron orbits [illustrated in \fig{fig:orbits}(a)]   are split in $\bk$-space with a differential area  $\delta S$ that is much less than their average area $S$\footnote{The expressions for $\delta S$ and $S$ in \q{deltaSvsS} are asymptotically valid for $E{\gg}m\alpha^2$.}:
\e{\delta S = 4\pi m \alpha k_E/\hbar\ll S =\pi k_E^2; \;\; k_{E}:=\f{\sqrt{2m E}}{\hbar}.\label{deltaSvsS}}
$S$ is equivalently the area of the constant-energy contour in the absence of spin-orbit coupling ($\alpha{=}0$).

@@

Our work presents a generalized rule to describe the Landau quantization of nearly-degenerate energy bands, for which  spin-orbit-split bands is but one example. By the \textit{near-degeneracy} condition, we generally mean that
the ratio of differential to average area is small: $|\delta S/S|{\ll}1$. Our proposed rule relies also on a second condition -- that $S$ is much larger than the inverse square of the magnetic length $l{=}\sqrt{\hbar/eB}$; this is the standard semiclassical condition in the Onsager-Lifshitz-Roth theory. By treating \textit{both} $|\delta S/S|$ and $1/l^2|S|$   as small parameters with finite ratio $l^2|\delta S|$, our perspective diverges from and generalizes \textit{single-parameter} semiclassical theories\cite{kohn_effham,blount_effham,rotheffham,wannier_fredkin,fischbeck_review,Mikitik_quantizationrule,topoferm,100p,gao_zero-field_2017}.

The inverse ratio $1/l^2|\delta S|$ measures the hybridization of nearly-degenerate orbits by the Zeeman interaction. In the regime $l^2|\delta S|{\gg}1$, the Onsager-Lifshitz{-Roth} quantization rules apply independently to  two concentric (but decoupled) orbits. In the opposite regime $l^2|\delta S|{\ll}1$, the spin-orbit splitting of energies is negligible, and we may then apply the known quantization rule for exactly-degenerate bands\cite{rotheffham,rothmag,topoferm,100p,Mikitik_quantizationrule}. Our unifying rule provides the smooth interpolation between these two known regimes. {Equivalently stated, we have extended the applicability of quantization rules to the intermediate regime ($l^2|\delta S|{\sim}1$) where the Zeeman interaction is comparable with the spin-orbit interaction; previously, such rules were known only when one interaction dominates over the other. This extension allows  to formulate semiclassical quantization rules for spin-orbit coupled materials in \textit{any} symmetry class -- with or without spatial inversion symmetry, with or without time-reversal symmetry. The unifying rule is presented in \s{sec:qtznrules} for nearly-degenerate bands with a spin-half (or pseudospin-half) degree of freedom. As elaborated in \s{sec:discussion}, the most general formulation of our rule is applicable  to any number of nearly-degenerate bands of any symmetry.}

In \s{sec:Rashba}, we demonstrate the utility of this rule for the Rashba{-Dresselhaus} 2DEG with an arbitrarily-oriented magnetic field. A sufficiently strong in-plane field not only moves the Dirac point but tilts the Dirac cone over -- into a type-II\cite{soluyanov_type-ii_2015, muechler_tilted_2016, bergholtz_topology_2015} Dirac point. In \s{sec:inplanezeeman}, we investigate the Landau-Zener tunneling dynamics that occur in the vicinity of a II-Dirac point; in particular, we will {demonstrate that our quantization rule can account for quantum tunneling between nearly-degenerate bands -- a phenomenon known as interband breakdown\cite{kaganov_coherent_1983,slutskin_dynamics_1968,AALG,100p}. We will also} critically evaluate a recent claim\cite{obrien_magnetic_2016} of perfect Klein tunneling at the energy of the Dirac point -- we find that such tunneling is never perfect with a proper account of the Zeeman interaction.

@@

In this section, we apply our rule to the Landau levels of the Rashba model [cf.\ \q{eq:Rashba-Hamiltonian}], for which analytic, closed-form solutions exist for verification\cite{bychkov_oscillatory_1984}.

For simplicity, we assume that the Pauli matrices in \q{eq:Rashba-Hamiltonian} correspond to spin: $s_i{=}\hbar \sigma_i/2$. Let us evaluate the effective Hamiltonian $\calh(\bk)$ along the cyclotron trajectory $\frako_0$ with radius $k$. In the energy basis of the Rashba Hamiltonian $H_R$ [cf.\ \q{eq:Rashba-Hamiltonian}], 
\e{\calh(\bk)=-\hbar\alpha k\tau_z{+}\f{g_{s\perp}}{2}\mu_{\sma{B}}B\tau_x {-} \f{eB\hbar}{2m} (\tau_x-I) \epsilon_{\alpha\beta} k_{\alpha}\nabla_{\beta}\theta,\label{rashbaeffham}}
where $\tau_i$ are a new set of Pauli matrices with $\tau_z{=}{+} 1$ and ${-}1$ corresponding to the outer and inner orbits, respectively. $e^{i\theta_{\bk}}{=}(ik_x{-}k_y)/k$ is the angle of the in-plane spin-splitting field $(-\hbar\alpha k_y,\hbar\alpha k_x,0)$ of $H_R$ [cf.\ \q{eq:Rashba-Hamiltonian}], which winds once over $\frako_0$; the spin of our basis vectors likewise winds, in parallel (or anti-parallel) alignment with this field (see Fig. (\ref{fig:orbits})(b)). Terms proportional to the rate of spin rotation ($\nabk \theta$) originate from the non-Abelian Berry connection. The first term of  $\calh$ corresponds to the spin-orbit-induced splitting of the energy-degenerate bands;
the remaining terms may be interpreted as $B$-induced deformations of the bands. The orbital moment term $B M_z$, which generally couples the two-band subspace (of interest) to all other bands, is absorbed into the Zeeman spin interaction with a phenomenological g-factor $g_{s\perp}$\footnote{For simplicity, we neglect here off-diagonal elements of $M_z$ in the basis of $S_z=\pm 1/2$, with $\vec{z}$ the direction of the field.}. It is well-known that the Zeeman term tends to polarize spin in the direction of $\bB$; remarkably, the non-Abelian Berry term  also has a polarizing effect, but with opposite sign. These two effects completely cancel for free electrons (with $g_{s\perp}=2$ and $m$ equal to the free-electron mass $m_0$), but not generically in band systems. 
%is absent in any simplified two-band model.

@@

In the weak-hybridization regime: $l^2\delta S \gg \pi|g_{s\perp} m/m_0{-}2|$, we may expand \q{eq:Rashba-lambda} as a Laurent series in $B$:
\e{\lambda_{\pm}= \pm \f{l^2\delta S}{2} +\pi \pm \f{\pi^2}{4}\f{(g_{s\perp}m/m_0-2)^2}{l^2\delta S} + O(B^2). \label{lambdasmallB}}
\q{eq:rule}, combined with the first two terms on the right-hand-side of \q{lambdasmallB}, may alternatively be derived from  the standard Onsager-Lifshitz rules for two independent orbits, with a common $\pi$ Berry-phase correction for each orbit\footnote{The intra-band Zeeman and orbital-moment couplings vanish due spacetime-inversion symmetry\cite{topoferm}.}: $l^2(S{\pm}\delta S)/2\pi {\in}\Z$. Higher-order terms in \q{lambdasmallB} reflect the field-induced hybridization of the spin-orbit-split orbits. Eq.\ (\ref{lambdasmallB}) has been partially obtained by Mineev for the Rashba model\cite{mineev_haas--van_2005}; he missed the second term in $(g_{s\perp}m/m_0-2)$ [cf.\ \q{lambdasmallB}], due to his neglect of the interband Berry connection.

@@

Our quantization rule [\q{eq:rule} with \q{eq:Rashba-lambda}] may be compared with the  exact Landau-level spectrum of the Peierls-substituted Hamiltonian $H_R(\bK)$ (plus the Zeeman coupling to spin).  The exact levels\cite{bychkov_oscillatory_1984} are given implicitly by
\e{
&n_{\pm}=\frac{l^2S}{2\pi}+\frac{l^2\delta S}{16\pi}\f{\delta S}{S}\lin 
&\pm\f1{2}\sqrt{\bigg(\frac{l^2\delta S}{8\pi}\f{\delta S}{S}\bigg)^2+(l^2\delta S)^2+\pi^2\bigg(\f{g_{s\perp}m}{m_0}-2\bigg)^2},\label{eq:Rashba-exact}}
with the functional forms of $\delta S(E)$ and $S(E)$ provided in \q{deltaSvsS};  $n_{+}{\ge} 1$ and $n_-{\ge} 0$ here are integer-valued labels for two distinct sets of levels. In the near-degenerate ($\delta S/S\ll 1$) and semiclassical ($1/l^2S \ll 1$) regime, with finite $l^2\delta S$, the two terms in \q{eq:Rashba-exact} involving $\delta S/S$  may be neglected, and \q{eq:Rashba-exact} reduces to our quantization rule with $\lambda_{\pm}$ given by \q{eq:Rashba-lambda}.

With the further inclusion of the Dresselhaus spin-orbit interaction in \q{eq:Rashba-Hamiltonian}, no closed-form analytic solution (of the Landau levels) is known. We may anyway verify the validity of our approximation scheme  -- by comparison with the Landau levels obtained from numerical exact diagonalization; this  is carried out in  \app{app:approximatevsexact}.

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{RZ.png}
\caption{(a) illustrates the cyclotron orbits of the Rashba 2DEG in a tilted magnetic field. The inset illustrates an enlarged view of near II-Dirac point. (b) illustrates the cyclotron orbit at the energy of the II-Dirac point; a Klein-tunneling electron will follow a trajectory that resembles the edge of a M\"obius strip. The orbit is drawn over the  spin-splitting field ($d_x,d_y$) of the degeneracy-lifting Hamiltonian $\delta\var$ [cf.\ \q{eq:RZ-Hamiltonian}]. The wavefunction along the inner (resp.\ outer) component of the M\"obius-strip orbit has its pseudospin aligned parallel (resp.\ anti-parallel) to the spin-splitting field. (c) schematically illustrates  Landau levels near the II-Dirac point (in the weak-hybridization regime). The dashed line represents the Onsager-Lifshitz-Roth quantization of the M\"obius-strip orbit [cf.\ Eq.\ (\ref{trefoilrule})], and the solid line incorporates the first-order correction [cf.\ Eq.\ (\ref{RZfirstorderE})] due to tunneling. (d) The black solid line is the plot of $\delta E_n$ [cf. Eq. (\ref{RZfirstorderE})], which is the energetic correction due to deviations from perfect Klein tunneling; for comparison, the red dashed line shows the numerically-exact deviations obtained from exact diagonalization. This comparison is made for the Landau levels closest to 0.45eV and with even Landau-level index $n$; for the parameters we have chosen ($m{=}0.076m_0$, $\alpha{=} 1.5\times10^{5}$cm/s, $Z_\parallel {=}$1meV and $g_{s\perp}=2$),  the II-Dirac point lies at 0.5eV. 
%Out-of-plane Zeeman coupling is negligibly weak.
\label{fig:RZ}}
\end{figure}

@@

As alluded to in \s{sec:qtznrules}, the adiabatic limit of two independent orbits does not exist if the two spin-orbit-split orbits exactly intersect at a II-Dirac point. To investigate this failure of adiabaticity, let us study the Landau levels at energies close to the II-Dirac point ($\epsilon_0$). We focus first on the regime where, on average (over $\frako_0$), the energy splitting $|\delta \var_+{-}\delta \var_-|$ [cf.\ \q{eq:RZ-Hamiltonian}] is much greater than $B_\perp|\calm_{\sma{+-}}|$. Equivalently, we are in the regime where $\hbar\alpha k_{\sma{E}}{\gg}\var_c^\perp$.\footnote{The orbit-averaged energy splitting is of order $\alpha k_{\sma{E}}$ for energies near the Dirac point. $B_\perp|\calm_{\sma{+-}}|$ is simply the Berry contribution to the Zeeman interaction, and has the form of the right-most term in \q{rashbaeffham}; this term is of order $\var_c$.} This implies that the split orbits do not appreciably hybridize -- except in the vicinity of the II-Dirac point where orbits (nearly) touch.  Near this touching point (where $t{=}0$), the dynamics of a spinor electron is described by linearizing $\H$ [cf.\ \q{calHRZ}] with respect to $t$:
\e{
\calh = -\hbar\alpha k_{E}\omega^{\sma{\perp}}_c t\,\sigma_x+(Z_\parallel-\alpha k_{E})\sigma_y-Z_\perp\sigma_z,\label{effhamIIDirac}
}
The eigenbasis of $\sigma_x$ corresponds to two diabatic levels with characteristic slope $v_d{:}{=}\hbar\alpha k_E\omega^{\sma{\perp}}_c$; near $t{=}0$, the diabatic levels anticross  due to the hybridization terms proportional to $\sigma_{y,z}$. Generally, the probability of tunneling across the hybridization-induced barrier is  $\rho^2{=}\exp(-2\pi\barmu)$, with $\barmu{=}{E_g}^2/{2v_d \hbar}$ and $E_g$ the energy gap at the center of the anticrossing\cite{wittig_landauzener_2005,lifshitz_e.m._quantum_1991}; in this context,
\e{\barmu=\frac{Z_\perp^2+(Z_\parallel-\alpha k_{E})^2}{2\hbar\alpha k_{E}\var^{\sma{\perp}}_c}, \label{mu1}}
with $2\hbar\alpha k_E$ the spin-orbit splitting at energy $E$ [cf.\ \fig{fig:orbits}(a)]. In the absence of the out-of-plane Zeeman coupling ($Z_\perp{=}0$), $\bar{\mu}$ would vanish at the energy of the II-Dirac point -- leading to unit-probability Klein tunneling, as was first proposed by Ref. \onlinecite{obrien_magnetic_2016} in a different model of the II-Dirac point. However, with a proper account of $Z_\perp$, we find instead that $\bar{\mu}$ has a nonvanishing minimum: $\bar{\mu}_{\sma{\text{min}}}{:}{=} (g_{s\perp} m/m_0)(Z_\perp/4\hbar\alpha k_{E})$, which is a product of the dimensionless effective mass and the ratio of the out-of-plane Zeeman splitting over the spin-orbit splitting. This minimum may be significant for semiconductor heterostructures (e.g.,  $g_{s\perp}m/m_0{\sim} 1$ in InAs heterostructures\cite{pakmehr_g-factor_2015}) and  for heavy-fermion systems.

@@

The appendices are organized as follows: in App. \ref{app:quantizationruleproof}, we prove the quantization rule of \qq{eq:rule}{berryconn}. The Landau levels derived from this quantization rule are then compared with numerically-exact Landau levels in App. \ref{app:approximatevsexact}, for the Rashba-Dresselhaus 2DEG in a magnetic field.  In App. \ref{sec:proofLLquasideg}, the existence of Landau-level quasidegeneracies is proven. In App. \ref{app:codimension}, we derive the symmetry-constrained codimensions of Landau-level quasidegeneracies, in all symmetry classes of cyclotron orbits.

%utilizing a mathematical result proved in App. \ref{app:makinggoffdiagonal}.

%The numerical diagonalization method is reviewed in App. \ref{sec:numerical}.
